---
title: "Does the Health of State Democracies Influence Satisfaction of Patients Using Government-Funded Healthcare Programs?"
author: "Ria Karthik"
description: "Comparing State Democracy Indices to Medicaid Patient Satisfaction"
output:
  distill::distill_article:
    self_contained: false
---


```{r, include = FALSE}
library(tidyverse)
library(broom)
```


## Introduction

I chose to study whether there is an association between the health of each US states' democracy and the satisfaction of patients receiving government-funded public healthcare. For the purposes of this study, I focused on comparing State Democratic Index values, and survey responses inquiring about satisfaction of care from adult patients receiving healthcare through Medicaid. I hypothesize that states with a lower democracy index are more likely to have lower satisfaction ratings from Medicaid patients on average. This hypothesis stems from the belief that political systems which lack proper democratic procedures are less likely to consider the needs of citizens and allocate funding and resources accordingly. 

This topic is especially important due to the increasing political polarization of the United States. Given that historically increases in political polarization are often associated with greater democratic backsliding, I believe it is important to study whether this backsliding and overall declining health of democracy in US states is affecting the quality of healthcare US citizens receive.  

## Data 

I have chosen two data sets, each ranking specific characteristics of states within the US. The first data set, health_quality looks at how specific government-funded health care programs for low income families qualitatively performs compared to other states. The data on quality of healthcare delivered by each state is drawn from ratings of Medicaid and the Children's Health Insurance Program (CHIP). For this study I will only be using data from patients 18 or older under Medicaid. These patients completed surveys answering whether they felt all their medical needs were meet on a 4-point scale from 1 being Never to 4 being Always under Medicaid. I plan to combine this information with my second dataset, the state_dem_inx which measures the State Democracy Index for each from 2000 to 2023. This data looks into each states average democratic performance quantitatively. I will be using the democracy_mcmc scores, to determined the average Electoral Democracy Scores over the years the study was conducted. 

Using these two datasets I plan to run a cross-sectional analysis to determine whether a states electoral democratic performance is a determinant in healthcare quality. Therefore, my explanatory variable will be the Electoral Democracy Score value of each state, as this is what I believe may influence healthcare quality. My outcome variable will be the average quality rating of each state's Medicaid program (on a 4-point scale from 1 being patient needs were "Never" met to 4 being "Always" met). 

To determine whether or not my hypothesis is correct, I will have to find the most relevant average publicly funded healthcare ratings for each state and compare it to the average state democracy index values (Electoral Democracy Scores) for each state. If states with lower index values tend to have lower healthcare quality ratings, this would affirm my hypothesis. If however, states with lower democratic index have higher healthcare ratings on average, this would disprove my theory. 

```{r, include = FALSE}
healthcare_quality <- read_csv("healthcare_quality.csv")
healthcare_quality

state_dem_ind <- read_csv("State_Dem_Index.csv") |>
  mutate(State = state) |> 
  group_by(State) |>
  summarize(mean_democracy_score = mean(democracy_mcmc, na.rm = TRUE))

state_dem_ind
```
```{r, echo = FALSE}
clean_healthcare_df <- healthcare_quality |>
  filter(Population == "Medicaid", 
         Domain == "Experience of Care", 
         `Reporting Program` != "Child Core Set", 
         `Rate Definition` == "Percentage who Selected They \"Always\" Got Needed Care, on a Four-Point Scale Ranging from \"Never\" to \"Always\": Age 18 and Older") |>
  select(c("State", "Domain", "Measure Name",  "Population", "Rate Definition", "State Rate")) |>
  mutate(`State Rate` = suppressWarnings( as.numeric(`State Rate`))) |>
  arrange(`State Rate`)

```

```{r, include = FALSE}
glimpse(clean_healthcare_df)
```

```{r, echo = FALSE}
clean_healthcare_df <- clean_healthcare_df |>
  arrange(State) |>
  mutate(Group = if_else(row_number() <= n()/2, "Group 1", "Group 2")) 


ggplot(data = clean_healthcare_df, 
       mapping = aes(x = reorder(State,`State Rate`, mean, decreasing = TRUE), 
                     y = `State Rate`)) +
       geom_col() + 
  # facet_wrap(~Group, scales = "free_x") +
  theme(axis.text.x = element_text(angle = 50, hjust = 1)) +
  labs(x = "State", y = "Average State Rating", title = "Average Patient Satisfaction Ratings by State on 4-Point Scale", caption = "Average State Rating measured on a 4-point scale of 1 being Never to 4 being Always")
```

The visual above shows average state rating on experience of care for patients 18 or older under Medicaid. The rating is based on the on whether or not patients felt they had received the care they needed on a 4 point scale (1 being "Never" to 4 being "Always"). The purpose of this graph is to get a general idea of which states tend to make patients feel more satisfied with the amount and quality of care they receive through Medicaid. Based on the graph, the state with the highest average satisfaction rating is Mississippi, and the state with the lowest available score is New York. The average state rating appears to be around 50. Using this we can further analyze if states that have higher democratic index values tend have higher or lower ratings of Medicaid. 

##  Results
```{r, echo = FALSE}
combined_df <- left_join(clean_healthcare_df, state_dem_ind, by = "State")
```
### Scatterplot
```{r, echo = FALSE}
ggplot(data = combined_df,
       mapping = aes(x = mean_democracy_score,
                     y = `State Rate`)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Mean State Democracy Index Score", y = "Average State Rating", title = "Comparison of Mean SDI Score to State Ratings")
```
The scatterplot displays a the relationship between mean state democratic index scores and the average satisfaction patients had with Medicare on a 4-point scale (1 being Never to 4 being Always) in each state. The plot displays a negative association between State Democratic Index values and average state ratings. Based on this graph alone it appears that states with less healthy democracies provide more satisfactory healthcare for patients using Medicare, however the linear regression below with provide further confirmation on whether or not such an association can be confirmed. 

### Linear Regression
```{r, echo = FALSE}
combined_df_lm <- lm(`State Rate` ~ mean_democracy_score, data = combined_df)

s <- summary(combined_df_lm)

coef_df <- as.data.frame(s$coefficients)

colnames(coef_df) <- c(
  "Estimate",
  "Standard Error",
  "t-value",
  "p-value"
)

rownames(coef_df) <- c("Intercept", "Mean Democracy Score")

knitr::kable(coef_df, digits = 2)
```

The data used in this linear regression was used a combined data set, containing the State Rankings of quality of care (on a four-point scale from "Always" to "Never") from patients above the age of 18 under Medicaid, and data on the mean Electoral State Democracy Scores for each state extracted from the State Democracy Index study. The aim of the linear regression, is to understand whether the democracy score is correlated with higher approval ratings for public healthcare recipients. 

The p-value of the Mean Democracy Score is nearly 0 which is less than the alpha value of 0.05, therefore fail to reject the null hypothesis (we reject the proposed alternative hypothesis). This p-value also indicates that the results of the regression are statistically significant. 

There also appears to be a negative association between ratings for healthcare provided to adults through Medicaid and the mean State Democracy Index Scores. For every point increase in state democracy index scores (moving 1 point towards a healthier democracy), the approval ratings for patients above the age of 18 under Medicaid go down by 4.25. The intercept value indicates that when the mean score of SDI values is 0 a state is considered neutral (not a healthy or unhealthy democracy), and the predicted approval rating was 52.16.That being said, the independent variable of this dataset (SDI values) was not collected through randomized sampling, therefore we can not interpret this relationship causally, instead we can only determine associations. 

## Conclusion

Based on this information, I believe that the association between the quality of publicly funded healthcare scores, and electoral democracy scores disprove my hypothesis that states with less healthy democracies will have lower approval ratings for publicly funded medical programs such as Medicaid. In fact the main estimated coefficient value was the opposite sign of what I had predicted indicating that states with lower electoral democracy ratings have higher average state ratings of quality of through Medicaid. 

There are however several potential confounders for these observations. One being the possibility that States being Red or Blue influences the quality and breadth of Medicaid coverage as well as the State Democratic Index values. Another possibility is that States with higher SDI values are more educated and therefore have higher expectations for quality of publicly funded healthcare programs. There are many other socioeconomic factors that could be influencing our observed outcomes which is why a randomized collection of the independent variable data is necessary. Additionally, this dataset was missing data from States like Utah which lowers the validity of this studies observations. 

If I were to run this study again with intent of being able to make causal inferences, I would randomize the experiment by measuring individual patient experiences. This study would randomly assign numbers to Medicaid recipients over the age of 18 and randomly select a large number to sample. The independent variable being the SDI value of the state they are from and the dependent variable being their personal satisfaction rating of Medicaid on the same 4-point scale. 


```{r print-code, ref.label=knitr::all_labels(), echo = TRUE, eval=FALSE}

```